---
title: "Racial/ethnic disparities in antibiotic use"
author: "Scott W. Olesen and Yonatan H. Grad"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(message=FALSE, echo=FALSE, cache=TRUE,
                      fig.path='fig/', dev=c('png', 'pdf'))
pdf.options(useDingbats=FALSE)

library(dplyr)
library(readr) # for reading the tsv's
library(magrittr) # for easier data piping
```

# Methods

## Variance estimation

For correct estimation of the variances in the complex survey, use the *survey* package:

```{r load_survey_package}
# for correct variance estimation
library(survey)

# tidying the survey functions
tidy.svyby = function(sb) {
  ci = confint(sb)
  df = data_frame(estimate=coef(sb), se=SE(sb),
                  cil=ci[,1], ciu=ci[,2])
  
  for (i in 1:(length(names(sb)) - 2)) {
    cn = names(sb)[i]
    df[[cn]] = sb[[cn]]
  }
  
  df
}

tidy.svystat = function(sm) {
  ci = confint(sm)
  data_frame(estimate=coef(sm), se=as.numeric(SE(sm)), cil=ci[,1], ciu=ci[,2])
}

citation('survey')
```

The [MEPS website](https://meps.ahrq.gov/survey_comp/hc_samplecodes_se.shtml) includes examples of using multiple software programs, including the *survey* package, for variance estimation.

## Survey data

The MEPS data for 2014/2015 are scraped and cleaned using a script in the `data/` directory. We exclude individuals with zero survey weight.

```{r load_meps_data}
# encoding and display strings for the two exposure variables
race_levels = c('white', 'other', 'Hispanic', 'black', 'Asian')
race_labels = c('Non-Hispanic white only', 'other/multiple race', 'Hispanic', 'Non-Hispanic black only', 'Non-Hispanic Asian only')

white_levels = c('any_white', 'no_white')
white_labels = c('any white race', 'no white race')

meps_dat = read_tsv('../data/meps.tsv') %>%
  # cast race as a factor for nice ordering in tables
  mutate(race = factor(race, levels=race_levels, labels=race_labels),
         white = factor(white, levels=white_levels, labels=white_labels))

# create the survey object
meps = svydesign(ids=~survey_psu,
                 strata=~survey_stratum,
                 weights=~survey_person_weight,
                 data=meps_dat, nest=TRUE) %>%
  subset(survey_person_weight > 0)
```

# Results

```{r compute_use}
# compute counts of antibiotics for each of these categories
key_levels = c('abx', 'penicillins', 'macrolides', 'other', 'sulfonamides', 'quinolones')
key_labels = c('all antibiotics', 'penicillins', 'macrolides', 'other antibiotics', 'sulfonamides', 'quinolones')
key_f = function(x) factor(x, levels=key_levels, labels=key_labels)

# antibiotics per capita across everyone
overall_mean = svymean(~n_abx, meps)

# antibiotics per capita, for each race/ethnicity and drug group
race_means = data_frame(key=key_levels, frmla=str_c('~n_', key)) %>%
  group_by(key) %>%
  do(tidy(svyby(as.formula(.$frmla), ~race, meps, svymean))) %>%
  ungroup() %>%
  mutate(key = key_f(key))

# and by white race
white_means = data_frame(key = key_levels, frmla = str_c('~n_', key)) %>%
  group_by(key) %>%
  do(tidy(svyby(as.formula(.$frmla), ~white, meps, svymean))) %>%
  ungroup() %>%
  mutate(key = key_f(key))
```

## Per capita use of all antibiotics

```{r overall_table_race, warning=FALSE}
# per capita use of all antibiotics, among everyone
overall_row = tidy(overall_mean) %>%
  mutate(race = 'everyone')

# per capita use of all antibiotics, by race/ethnicity
race_rows = race_means %>%
  filter(key == 'all antibiotics')

bind_rows(overall_row, race_rows) %>%
  select(`Race/ethnicity`=race, `Annual fills per 1,000 people`=estimate,
         `Lower 95% CI`=cil, `Upper 95% CI`=ciu) %>%
  mutate_if(is.numeric, funs(round(. * 1000))) %>%
  kable(caption='Reported antibiotic use by race/ethnicity category, MEPS 2014/2015')
```

```{r overall_table_white}
white_means %>%
  filter(key == 'all antibiotics') %>%
  select(`Race`=white, `Annual fills per 1,000 people`=estimate,
         `Lower 95% CI`=cil, `Upper 95% CI`=ciu) %>%
  mutate_if(is.numeric, funs(round(. * 1000))) %>%
  kable(caption='Reported antibiotic use by white race, MEPS 2014/2015')
```

## Per capita use by drug

```{r use_by_race, fig.width=3.54, fig.height=3.5}
my_palette = c("#FFFFFF", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

race_plot = race_means %>%
  mutate_if(is.numeric, funs(.*1000)) %>%
  mutate(key=fct_reorder(factor(key), -estimate)) %>%
  ggplot(aes(key, estimate, fill=race)) +
  geom_col(position='dodge', color='black') +
  geom_errorbar(aes(ymin=cil, ymax=ciu), position=position_dodge(width=0.9), width=0.5) +
  scale_fill_manual(values=my_palette) +
  xlab('') +
  ylab('reported annual fills per 1,000 people') +
  ylim(0, 500) +
  theme(panel.background=element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(color='gray'),
        panel.grid.minor.y = element_line(color='gray'),
        axis.line.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.title = element_blank(),
        legend.background = element_rect(color='black'),
        legend.position=c(0.8, 0.8),
        legend.margin = margin(0, 5, 4, 5),
        legend.text = element_text(size=8),
        legend.key.size = unit(4, 'mm'),
        axis.text.x = element_text(size=8, color='black', angle=30, hjust=1.0, vjust=1.25),
        axis.text.y = element_text(size=7, color='black'),
        axis.title.y = element_text(size=9)
  )

race_plot
```

```{r use_by_white, fig.width=3.54, fig.height=3.5}
white_plot = white_means %>%
  mutate_if(is.numeric, funs(.*1000)) %>%
  mutate(key=fct_reorder(factor(key), -estimate)) %>%
  ggplot(aes(key, estimate, fill=white)) +
  geom_col(position='dodge', color='black') +
  geom_errorbar(aes(ymin=cil, ymax=ciu), position=position_dodge(width=0.9), width=0.5) +
  scale_fill_manual(values=c('white', 'black')) +
  xlab('') +
  ylab('') +
  ylim(0, 500) +
  theme(panel.background=element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(color='gray'),
        panel.grid.minor.y = element_line(color='gray'),
        axis.line.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.title = element_blank(),
        legend.background = element_rect(color='black'),
        legend.position=c(0.7, 0.865),
        legend.margin = margin(0, 5, 4, 5),
        legend.text = element_text(size=8),
        legend.key.size = unit(4, 'mm'),
        axis.text.x = element_text(size=8, color='black', angle=30, hjust=1.0, vjust=1.25),
        axis.text.y = element_text(size=7, color='black'),
        axis.title.y = element_text(size=9)
  )

white_plot +
  ylab('reported annual fills per 1,000 people')
```

```{r combined_plot, fig.width=7.5, fig.height=3.5, eval=FALSE}
ggpubr::ggarrange(race_plot, white_plot, labels=c('a', 'b'), widths=c(1, 3/5))
```

## Race/ethnic disparities in use

```{r compute_ratios}
ratio = function(num_mean, num_sd, den_mean, den_sd, n_trials=1e6, alpha=0.05) {
  # draw random deviates to simulate the random ratios
  numerator_deviates = rnorm(n_trials, num_mean, num_sd)
  denominator_deviates = rnorm(n_trials, den_mean, den_sd)
  ratios = numerator_deviates / denominator_deviates
  
  # compute the confidence interval from those random ratios
  cis = quantile(ratios, c(alpha / 2, 1 - alpha / 2))
  
  data_frame(estimate = num_mean / den_mean,
             se = sd(ratios),
             conf.low = cis[1],
             conf.high = cis[2])
}

race_ratios = race_means %>%
  filter(race != 'Non-Hispanic white only') %>%
  left_join(filter(race_means, race=='Non-Hispanic white only'), by='key', suffix=c('', '.ref')) %>%
  group_by(key, race) %>%
  do(ratio(.$estimate.ref, .$se.ref, .$estimate, .$se)) %>%
  ungroup()

white_ratios = white_means %>%
  filter(white != 'any white race') %>%
  left_join(filter(white_means, white=='any white race'), by='key', suffix=c('', '.ref')) %>%
  group_by(key, white) %>%
  do(ratio(.$estimate.ref, .$se.ref, .$estimate, .$se)) %>%
  ungroup()
```

```{r ratio_tables}
race_ratios %>%
  select(`Antibiotic`=key, `Race`=race, `ratio (NHWO vs. this race)`=estimate,
         `lower 95% CI`=conf.low, `upper 95% CI`=conf.high) %>%
  mutate_if(is.numeric, function(x) round(x, 3)) %>%
  kable(caption='Ratios of reported fill rates by categorical race/ethnicity, MEPS 2014/2015')

white_ratios %>%
  select(`Antibiotic`=key, `ratio (any white race vs. no white race)`=estimate,
         `lower 95% CI`=conf.low, `upper 95% CI`=conf.high) %>%
  mutate_if(is.numeric, function(x) round(x, 3)) %>%
  kable(caption='Ratios of reported fill rates by white race, MEPS 2014/2015')
```
