---
title: "Racial/ethnic disparities in antibiotic use"
author: "Scott W. Olesen and Yonatan H. Grad"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(message=FALSE, echo=FALSE, cache=TRUE,
                      fig.path='fig/', dev=c('png', 'pdf'))
pdf.options(useDingbats=FALSE)
```

# Methods

## Variance estimation

For correct estimation of the variances in the complex survey, use the *survey* package:

```{r load_survey_package}
# for correct variance estimation
library(survey)

# tidying the survey functions
tidy.svyby = function(sb) {
  ci = confint(sb)
  df = data_frame(estimate=coef(sb), se=SE(sb),
                  cil=ci[,1], ciu=ci[,2])
  
  for (i in 1:(length(names(sb)) - 2)) {
    cn = names(sb)[i]
    df[[cn]] = sb[[cn]]
  }
  
  df
}

tidy.svystat = function(sm) {
  ci = confint(sm)
  data_frame(estimate=coef(sm), se=as.numeric(SE(sm)), cil=ci[,1], ciu=ci[,2])
}

citation('survey')
```

## Survey data

The MEPS data for 2014/2015 are scraped and cleaned using a script in the `data/` directory. We exclude individuals with zero survey weight and "inapplicable" ages (coded as `-1`).

```{r load_meps_data}
races = c('white', 'black', 'Hispanic', 'other')
race_labels = c('white', 'black', 'Hispanic', 'other race')

meps_dat = read_tsv('../data/meps.tsv') %>%
  # cast race as a factor for nice ordering in tables
  mutate(race=factor(race, levels=races, labels=race_labels)) %>%
  # dichotomous white-race variable for Poisson regressions
  mutate(white=as.integer(race=='white')) %>%
  # age groups for Poisson regressions
  mutate_at(vars(age), as.integer) %>%
  mutate(age_group = case_when(
    .$age == -1 ~ 'inapplicable',
    between(.$age, 0, 9) ~ '0-9',
    between(.$age, 10, 20) ~ '10-20',
    between(.$age, 21, 64) ~ '21-64',
    .$age >= 65 ~ '65+'
  ))

meps = svydesign(ids=~survey_psu,
                 strata=~survey_stratum,
                 weights=~survey_person_weight,
                 data=meps_dat, nest=TRUE) %>%
  subset(survey_person_weight > 0 & age_group != 'inapplicable')
```

# Results

```{r compute_race_category}
# compute counts of antibiotics for each of these categories
key_levels = c('abx', 'penicillins', 'macrolides', 'other', 'sulfonamides', 'quinolones')
key_labels = c('all antibiotics', 'penicillins', 'macrolides', 'other antibiotics', 'sulfonamides', 'quinolones')
key_f = function(x) factor(x, levels=key_levels, labels=key_labels)

# antibiotics per capita across everyone
overall_mean = svymean(~n_abx, meps)

# antibiotics per capita, for each race/ethnicity and drug group
race_means = lapply(key_levels, function(abx) {
  svyby(as.formula(str_c('~n_', abx)), ~race, meps, svymean) %>%
    tidy() %>%
    mutate(key=abx)
}) %>%
  bind_rows() %>%
  mutate(key=key_f(key))
```

## Per capita use of all antibiotics

```{r overall_table}
# per capita use of all antibiotics, among everyone
overall_row = overall_mean %>%
  tidy() %>%
  mutate(race='everyone')

# per capita use of all antibiotics, by race/ethnicity
race_rows = race_means %>%
  filter(key=='all antibiotics')
  
bind_rows(overall_row, race_rows) %>%
  select(`Race/ethnicity`=race, `Annual fills per 1,000 people`=estimate,
         `Lower 95% CI`=cil, `Upper 95% CI`=ciu) %>%
  mutate_if(is.numeric, funs(round(. * 1000))) %>%
  kable(caption='Per capita antibiotic use, MEPS 2014/2015')
```

## Per capita use of drug groups

```{r use_by_race, fig.width=3.54, fig.height=3.5}
my_palette = c("#FFFFFF", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

race_means %>%
  mutate_if(is.numeric, funs(.*1000)) %>%
  mutate(key=fct_reorder(factor(key), -estimate)) %>%
  ggplot(aes(key, estimate, fill=race)) +
  geom_col(position='dodge', color='black') +
  geom_errorbar(aes(ymin=cil, ymax=ciu), position=position_dodge(width=0.9), width=0.5) +
  scale_fill_manual(values=my_palette) +
  xlab('') +
  ylab('annual fills per 1,000 people') +
  theme(panel.background=element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(color='gray'),
        panel.grid.minor.y = element_line(color='gray'),
        axis.line.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.title = element_blank(),
        legend.background = element_rect(color='black'),
        legend.position=c(0.8, 0.8),
        legend.margin = margin(0, 5, 4, 5),
        legend.text = element_text(size=8),
        legend.key.size = unit(4, 'mm'),
        axis.text.x = element_text(size=8, color='black', angle=30, hjust=1.0, vjust=1.25),
        axis.text.y = element_text(size=7, color='black'),
        axis.title.y = element_text(size=9)
  )
```

## Race/ethnic disparities in use

For each drug group, we run a Poisson regression predicting per capita antibiotic use from race (white or not), correcting for covariates (age group, sex, US Census region). This regression excludes people without a well-defined Census region.

```{r race_models}
race_models = lapply(key_levels, function(abx) {
  svyglm(as.formula(str_interp("n_${abx} ~ white + sex + age_group + region")),
         subset(meps,  region != 'no_region'), family='quasipoisson') %>%
    tidy() %>%
    mutate(key=abx)
}) %>%
  bind_rows() %>%
  mutate(key=key_f(key))
```

For each model, we report the fold-difference in per-capita use between whites and non-whites. The $p$-values are for a fold difference not equal to 1.

```{r race_models_table}
race_models %>%
  filter(term=='white') %>%
  mutate(ci=1.96*std.error, cil=estimate-ci, ciu=estimate+ci) %>%
  # svyglm returns the log risk ratio; exponentiate (and round) that
  # to get the risk ratio (i.e., fold difference)
  mutate_at(vars(estimate, cil, ciu), funs(round(exp(.), 2))) %>%
  select(`Drug group`=key, `Fold difference`=estimate,
         `Lower 95% CI`=cil, `Upper 95% CI`=ciu, p.value) %>%
  mutate(p.value=prettyNum(p.value, digits=3)) %>% 
  kable(caption='Fold-differences in per capita use, whites vs. non-whites; MEPS 2014/2015')
```
